Transition path sampling based calculation of free energies of enzymatic reactions

### 2022-04-07 tex/manuscript.tex
added some more figures

TPS in particular is an attractive method since it samples reactive trajectories
in which the rare barrier-crossing events are guaranteed to occur and
it does not require a priori knowledge of the reaction coordinate which can be 
a complicated entity in large biological systems. TPS has been successfully 
applied to study problems such as enzyme catalysis, protein folding, etc. 
The TPS ensemble which is used to obtain a unique statistical definition for the reaction coordinate
and transition state of a reaction consists of so called reactive trajectories which 
link the stable reactant and product states. In this Monte-Carlo procedure in the trajecotry space
rejected moves correspond to non-reactive trajectories. But these trajectories are 
required to obtain the equilibrium distribution function.

@@

In this work we will focus on the implementation and application of the 
free energy based calculations within the TPS method on two enzymatic reactions: 
One example of the TPS based free energies calculations on real 
protein systems \cite{Silveira21JPhysChemB125p2018} Human Mat2a enzyme and 
plasmodium vivax adenosine deaminase \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
%Transition path sampling is an importance sampling in the trajectory space.
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

%------------------------------------------------------------------------------ 
\subsection{Human MethionineAdenosyltransferase 2A}
%------------------------------------------------------------------------------ 
The SN2 reaction catalyzed by the human Mat2a enzyme leads to the production of 
S-Adenosylmethionine compound which is part of the ATP cycle \cite{Niland21Biochem60p791}
\begin{figure}[ht!]
\includegraphics[scale=0.5]{figures/comm-80-mat2a.pdf}
\end{figure}

>>>>

TPS was developed as a reaction-coordinate free, statistical method 
which samples trajectories that connect the stable reactant and product states 
guaranteeing the occurence of the rare barrier crossing events. These trajectories 
are sampled using a Monte carlo procedure leading to an ensemble of transition
pathways that are real dynamical trajectories which contain information regarding the 
mechanism and kinetics of the reaction.

TPS in particular is an attractive method since it samples reactive trajectories
in which the rare barrier-crossing events are guaranteed to occur and
it does not require a priori knowledge of the reaction coordinate which can be 
a complicated entity in large biological systems. TPS has been successfully 
applied to study problems such as enzyme catalysis, protein folding, etc. 
 In this Monte-Carlo procedure in the trajecotry space
rejected moves correspond to non-reactive trajectories. But these trajectories are 
required to obtain the equilibrium distribution function.

@@

In this manuscript we will focus on the implementation and application of the 
free energy based calculations within the TPS method on reactions catalyzed by two systems:
the human MAT2A enzyme and the plasmodium vivax adenosine deaminase enzyme. \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
These enzymes were chosen not only because of their importance in biochemical pathways and
but also because of the availability of experimental results with which comparisons can 
be drawn.

One example of the TPS based free energies calculations on real 
protein systems \cite{Silveira21JPhysChemB125p2018} Human Mat2a enzyme and 
plasmodium vivax adenosine deaminase 
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
Here we briefly describe the equations describing the TPS method for the purpose of
introducing notations as well as for the sake of being complete. 
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
The SN2 reaction catalyzed by the human Mat2a enzyme leads to the production of 
S-Adenosylmethionine compound which is part of the ATP cycle \cite{Niland21Biochem60p791}
\begin{figure}[ht!]
\includegraphics[scale=0.25]{figures/mat2a-transition-state.png}
\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=0.5]{figures/comm-80-mat2a.pdf}
\end{figure}
